A Flexible Machine Learning-Aware Architecture\\ for Future WLANs

### 2020-01-20 LaTeX files/flexible_ml_architecture_wlans.tex
Progress on the reviewed version of the manuscript + Code NN training + Update README

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
Lots of hopes have been placed \textcolor{orange}{on} \textcolor{blue}{machine learning (ML)} as a key enabler of future wireless networks. By taking advantage of the large volumes of data generated by networks, ML is expected to deal with the ever-increasing complexity of networking problems. Unfortunately, current \textcolor{orange}{networks} are not yet prepared \textcolor{orange}{to support} the ensuing requirements of ML-based applications \textcolor{orange}{in terms of} data collection, processing, and output distribution. This article points out the architectural requirements that are needed to pervasively include ML as part of future wireless networks operation. \textcolor{orange}{Specifically, we look into \textcolor{blue}{wireless local area networks (WLANs)}, which, due to their nature, can be found in multiple forms, ranging from cloud-based to edge-computing-like deployments. In particular,} we propose to adopt the International Telecommunications Union (ITU) unified architecture for 5G and beyond. Based on the ITU's architecture, we provide insights on the main requirements and the major challenges of introducing ML to the multiple modalities of WLANs. %\textcolor{blue}{Finally, numerical results show the superiority of ML-based methods enabled by the architecture in front of currently adopted approaches.}
\end{abstract}

@@

ML is meant to empower a computational system for learning automatically, based on experience, so that future situations can be properly managed without having been programmed explicitly. Concerning wireless communications, there is a huge amount of unexploited data generated at both infrastructure and user levels, which could be extremely useful \textcolor{orange}{\st{to ML mechanisms}} for learning complex patterns\textcolor{orange}{, which may allow to improve network performance}. For instance, the behavior of users in a given network-oriented service can be predicted through ML, based on the information from previous sessions. \textcolor{orange}{Based on these predictions, network resources can be properly accommodated} in future sessions, thus leading to an optimized solution.

Unfortunately, the potential benefits of ML \textcolor{orange}{\st{solutions}} for real networks are currently limited by the existing infrastructure, which is not yet prepared to accommodate ML-oriented tasks such as data collection, processing, and output distribution. Instead, current networking systems are typically meant for delivering content, without taking into account the underlying characteristics of the processes that generate it.

@@

Apart from the ITU-T initiatives, other important standardization bodies such as the 3rd Generation Partnership Project (3GPP) or the European Telecommunications Standards Institute (ETSI) are currently working on the integration of data analytics to network functions. The 3GPP contemplates AI as one of the priority topics for shaping its upcoming release (Release 17), and architectural requirements are currently under study \cite{3gpp2019study}. Furthermore, we highlight the ETSI groups on Experiential Networked Intelligence (ENI) and Zero touch network and Service Management (ZSM), which actively study the integration of AI to networks \cite{etsi2019architecture}. Unlike the ITU's unified architecture, most of the work held by the 3GPP and the ETSI focuses on centralized data collection and data analytics solutions. Nevertheless, we understand that the works in \cite{itu2019architecture, 3gpp2019study, etsi2019architecture} are complementary to each other.
% Alternative architecture: \cite{nekovee2019mobile}

@@

% Overview ML with formal definitions
In order to meet the abovementioned strict requirements, not only a technological innovation is expected (e.g., use of more spectrum or improve multiple-antennas technologies), but a paradigm shift is necessary when designing solutions for network planning, operation, and management. In particular, intelligent wireless networks need to be empowered with cognitive and context-aware capabilities, which may require additional infrastructure such as environmental sensors and cameras. To that end, ML is expected to be important during the lifetime of 5G and will become pervasive - as included from the beginning in their conception - in 6G networks.

@@

% 1 - Architectural proposals	
Apart from the specific ML solutions to problems in communications, some efforts have been made towards enabling AI-aware networking in more general terms. In this regard, several architectural proposals have been provided so far \cite{bi2015wireless,chih2017big,wang2018machine}. Most of the referenced works agree in the necessary steps for enabling big data analytics in cellular deployments\textcolor{orange}{\st{, which roughly are}}: (1) data collection, (2) data preparation, (3) data analysis, and (4) decision making. Nevertheless, none of these works provide architectural guidelines to introduce ML to wireless networks. In contrast, the ITU's architecture looks deeper into the ML operation and targets the actual procedures involving information gathering, processing, and communication. In addition to the architectural guidelines, the ITU-T has provided a data handling framework for ML-aware networks \cite{itu2019data}, which defines processes concerning data collection, processing, and output distribution.

@@

\subsubsection{Cloud-Based User Association and Handover}
Most of the current user association and handover procedures held in WLANs typically rely on strongest signal first (SSF) mechanisms. This might be problematic in terms of load balancing \textcolor{orange}{and can potentially lead} to severe performance degradation \textcolor{orange}{in} saturated \textcolor{blue}{basic service sets (BSSs)}. By introducing ML, it is possible to handle contextual information such as the traffic load \textcolor{orange}{\st{at the APs}}, which can \textcolor{orange}{be useful for} decision-making. Furthermore, mobility pattern prediction and predicted user requirements can be included in the system, thus empowering the association and handover mechanisms with insightful information. %\textcolor{orange}{Predictions can be provided, for instance, by Deep Neural Networks (DNN).}

@@

\subsubsection{Reinforcement Learning-Based Spatial Reuse} 
\textcolor{orange}{\st{In contrast to centralized and cooperative ML approaches, local optimization is required for independent WLANs. In this regard,}} Spatial reuse aims to improve channel utilization by means of sensitivity adjustment mechanisms. However, selecting the best sensitivity threshold is not trivial, given the complex spatial interactions that occur among WLANs and the high action space. \textcolor{orange}{To that} purpose, reinforcement learning can be introduced to improve spectral efficiency in a decentralized manner, so that each WLAN learns based on experience.

@@

The FG-ML5G was created in November 2017 by its parent group, the ITU-T Study Group 13, with the aim of studying the integration of ML mechanisms into future networks. This includes the definition of interfaces, protocols, data formats, and architectures. During its lifetime, the FG-ML5G has released several reports and contributions. Among them, we highlight the ITU's logical interoperable architecture for future networks \cite{itu2019architecture}, which defines an ML overlay that operates on the top of any unspecified underlay network technology (e.g., 3GPP, EdgeX, IEEE 802.11). The ITU's architecture aims to fulfill a set of technology-agnostic high-level requirements by the utilization of ML. For instance, the architecture must be able to support multiple types of data, thus taking advantage of heterogeneous data sources.

Figure \ref{fig:itu_ml_architecture} shows the elements that compose the ML overlay (management subsystem, multi-level ML pipeline, and closed-loop subsystem), which are further described in the following subsections. Based on this \textcolor{orange}{standard} overlay, ML applications can be instantiated in the logical entities (represented by white boxes).\textcolor{orange}{\st{, which are managed in a standard manner.}} %Communication among entities is done through input and output interfaces.

\subsection{Management Subsystem} 
The management subsystem is in charge of the deployment and the orchestration of the ML services \textcolor{orange}{that operate} in the underlying network. To that purpose, the \textcolor{blue}{machine learning function orchestrator (MLFO)} entity is defined. \textcolor{blue}{The MLFO is first instantiated by a declarative intent that uses a meta language, which specifies the ML use case to be applied, including initialization, policies and constraints. Then, the MLFO initializes the elements of the multi-level ML pipeline in the corresponding devices and monitors their operation during execution.}

@@

Figure \ref{fig:overview_learning_approaches} illustrates the enterprise and residential-like deployments, as well as a set of IEEE 802.11 mechanisms (summarized in Table \ref{tab:opportunities_amendments}) that can \textcolor{orange}{facilitate the adoption of the ML-based architecture in WLANs.} These mechanisms can be used for data collection, training, and output distribution.

@@

\textcolor{red}{TODO: convert table II into bullets.}

\begin{table*}[t!]
	\caption{Features supported by 802.11 amendments that can enable the introduction of ML methods to WLANs.}
	\label{tab:opportunities_amendments}
	\centering
	\begin{tabular}{|p{.14\textwidth}|p{.1\textwidth}|p{.6\textwidth}|}
		\hline
		\textbf{Feature} & \textbf{Amendments} & \textbf{Opportunities for ML application} \\\hline
		Information gathering & 802.11k/r/v & A given ML mechanism can use information about the network topology or RF measurements to infer the behavior of other devices or to extract important environmental characteristics.\\\hline
		Interoperability & 802.11f/u & Interoperability with other networks can be used to perform coordinated operations (e.g., scheduling, resource allocation). Besides, inter-AP communication procedures can enable centralized/coordinated mechanisms (e.g., federated learning). \\\hline
		Security & 802.11w & ML mechanisms can use management frames that are protected so that a higher level of security is granted.\\\hline
		Validation & 802.11t & Performance evaluation in WLANs through test metrics can be of great utility to define optimization goals within the ML  operation.\\\hline
	\end{tabular}
\end{table*}

%%% CHALLENGES IN IEEE 802.11 WLANs
\subsection{Challenges in IEEE 802.11 WLANs}
\label{section:ieee_80211_wlans}
The application of ML methods in WLANs is tightly tied to the technological challenges posed by these types of networks. The major challenges encountered in wireless communications stand for fast data expiry and lack of resources for data handling (e.g., storage, computation, information exchange). Regarding Wi-Fi networks, we find the following challenges:
\begin{enumerate}
	\item \textbf{Network dynamics:} channel fluctuations (e.g., due to multipath fading), users mobility and varying traffic needs entail a big challenge to ML applications. As a result of network dynamics, ML methods need to constantly adapt to the environment, which may be achieved through continuous re-training.
	\item \textbf{Limited communication resources:} since Wi-Fi operates under unlicensed bands, resources are scarce and shared. Thus, any potential communication required by a certain ML mechanism (as for distributed learning) may fail or be delayed if the medium is congested. As a result, the ML operation must be robust and resilient enough in order to react to potential communication issues.
	\item \textbf{Limited computation and storage resources:} computation and storage resources may also be limited in WLANs, especially in residential-like deployments. Therefore, the ML operation should include computation-efficient procedures. Another implication of limited resources lies in the availability of information to be used by ML algorithms, especially for online learning methods.
	\item \textbf{Adversarial environment:} in many cases, Wi-Fi deployments are chaotic in the sense that many overlapping \textcolor{orange}{BSSs} coexist without any kind of cooperation \textcolor{orange}{\st{(e.g., residential buildings)}}. This is a particularly interesting challenge for ML \textcolor{orange}{methods}, which may lead to adversarial settings where different agents compete for the same resources. Moreover, multi-vendor devices may implement different ML mechanisms that may lead to clashing interests.
	\item \textbf{Legacy devices:} \textcolor{orange}{similarly} to the previous case, ML-empowered WLANs may coexist with other legacy networks that do not perform any intelligent operation. It is then required for ML methods to be aware of those legacy devices, so that \textcolor{orange}{unfair} situations are avoided.
\end{enumerate}

@@

In edge-oriented mechanisms, the ML operation is mainly ruled by edge devices (e.g., APs and/or STAs), which, contrary to the cloud approach, typically lack powerful computation and storage resources. In consequence, edge-oriented mechanisms may only allow using simple and light-weight computing ML algorithms. Nevertheless, edge servers can be placed for providing more powerful solutions in a timely manner. The edge-oriented approach is useful for real-time ML applications that manage local (and even highly-varying) information.

Besides cloud and edge-oriented settings, we may distinguish between methods based on \textcolor{orange}{their cooperation degree}. In cooperative approaches, nodes interact among them for the sake of conducting the learning operation in a joint manner (e.g., use a shared reward). With this aim, reliable and timely connections among learners are often required. In this regard, \cite{lin2017deep} showed the role of communications on speeding up a distributed training procedure over a set of nodes in a network. Alternatively, for the non-cooperative case, the learning operation may lead to adversarial settings, especially since WLANs share resources such as frequency channels.

@@

Specifically, the training procedure consists of the following steps (shown \textcolor{orange}{in} red):
\begin{enumerate}
	\item \textbf{Data collection:} the edge server collects information of different \textcolor{orange}{kind} from either APs and STAs, \textcolor{orange}{\st{which can provide contextual data}} such as user information (e.g., current location), performance (e.g., throughput, delay), application data (e.g., traffic load), or channel status reports (e.g., the power received from interfering nodes). This information can be used either for training or feeding auxiliary algorithms that help the main AP association procedure (e.g., predict user behavior).
	\item \textbf{ML input preparation:} the data collected at the cloud is prepared by the pre-processor so that the ML method can properly manage it. For instance, in case of applying a multiple linear regression, the input user information needs to be converted into normalized features (e.g., convert the throughput given in Mbps into a scalar between 0 and 1).
	\item \textbf{ML method application:} on applying the ML method, policies are also taken into account for generating the final output result. In this case, a given AP may have specific constraints regarding its occupation or the maximum number of associated STAs. The policies are strongly tied to the capabilities of the devices or the existing regulations (e.g., frequency channels available due to antenna capabilities, maximum regulated transmission power).
	\item \textbf{Output distribution:} once the ML method generates the output (i.e., the predicted function for new (re)associations), it is distributed throughout the sink edge servers, which are then prepared to give quick response to new cases.
\end{enumerate}

@@

% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.
% For figures, \caption should occur after the \includegraphics.
% Note that IEEEtran v1.7 and later has special internal code that
% is designed to preserve the operation of \label within \caption
% even when the captionsoff option is in effect. However, because
% of issues like this, it may be the safest practice to put all your
% \label just after \caption rather than within \caption{}.
%
% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
% option should be used if it is desired that the figures are to be
% displayed while in draft mode.
%
%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{myfigure}
% where an .eps filename suffix will be assumed under latex, 
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
%\caption{Simulation results for the network.}
%\label{fig_sim}
%\end{figure}

% Note that the IEEE typically puts floats only at the top, even when this
% results in a large percentage of a column being occupied by floats.

% An example of a double column floating figure using two subfigures.
% (The subfig.sty package must be loaded for this to work.)
% The subfigure \label commands are set within each subfloat command,
% and the \label for the overall figure must come after \caption.
% \hfil is used as a separator to get equal spacing.
% Watch out that the combined width of all the subfigures on a 
% line do not exceed the text width or a line break will occur.
%
%\begin{figure*}[!t]
%\centering
%\subfloat[Case I]{\includegraphics[width=2.5in]{box}%
%\label{fig_first_case}}
%\hfil
%\subfloat[Case II]{\includegraphics[width=2.5in]{box}%
%\label{fig_second_case}}
%\caption{Simulation results for the network.}
%\label{fig_sim}
%\end{figure*}
%
% Note that often IEEE papers with subfigures do not employ subfigure
% captions (using the optional argument to \subfloat[]), but instead will
% reference/describe all of them (a), (b), etc., within the main caption.
% Be aware that for subfig.sty to generate the (a), (b), etc., subfigure
% labels, the optional argument to \subfloat must be present. If a
% subcaption is not desired, just leave its contents blank,
% e.g., \subfloat[].

% An example of a floating table. Note that, for IEEE style tables, the
% \caption command should come BEFORE the table and, given that table
% captions serve much like titles, are usually capitalized except for words
% such as a, an, and, as, at, but, by, for, in, nor, of, on, or, the, to
% and up, which are usually not capitalized unless they are the first or
% last word of the caption. Table text will default to \footnotesize as
% the IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
%\begin{table}[!t]
%% increase table row spacing, adjust to taste
%\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
%\caption{An Example of a Table}
%\label{table_example}
%\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
%\begin{tabular}{|c||c|}
%\hline
%One & Two\\
%\hline
%Three & Four\\
%\hline
%\end{tabular}
%\end{table}

% Note that the IEEE does not put floats in the very first column
% - or typically anywhere on the first page for that matter. Also,
% in-text middle ("here") positioning is typically not used, but it
% is allowed and encouraged for Computer Society conferences (but
% not Computer Society journals). Most IEEE journals/conferences use
% top floats exclusively. 
% Note that, LaTeX2e, unlike IEEE journals/conferences, places
% footnotes above bottom floats. This can be corrected via the
% \fnbelowfloat command of the stfloats package.

@@

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://mirror.ctan.org/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
\bibliographystyle{IEEEtran}
\bibliography{bibliography}
%\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
%\begin{thebibliography}{1}
%
%\bibitem{IEEEhowto:kopka}
%H.~Kopka and P.~W. Daly, \emph{A Guide to \LaTeX}, 3rd~ed.\hskip 1em plus
%  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
%
%\end{thebibliography}

% biography section
% 
% If you have an EPS/PDF photo (graphicx package needed) extra braces are
% needed around the contents of the optional argument to biography to prevent
% the LaTeX parser from getting confused when it sees the complicated
% \includegraphics command within an optional argument. (You could create
% your own custom macro containing the \includegraphics command to make things
% simpler here.)
%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
% or if you just want to reserve a space for a photo:

@@

% insert where needed to balance the two columns on the last page with
% biographies
%\newpage

@@

>>>>

\usepackage{url}

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
Lots of hopes have been placed \textcolor{orange}{on} \textcolor{blue}{machine learning (ML)} as a key enabler of future wireless networks. By taking advantage of the large volumes of data generated by networks, ML is expected to deal with the ever-increasing complexity of networking problems. Unfortunately, current \textcolor{orange}{networks} are not yet prepared \textcolor{orange}{to support} the ensuing requirements of ML-based applications \textcolor{orange}{in terms of} data collection, processing, and output distribution. This article points out the architectural requirements that are needed to pervasively include ML as part of future wireless networks operation. \textcolor{orange}{Specifically, we look into \textcolor{blue}{wireless local area networks (WLANs)}, which, due to their nature, can be found in multiple forms, ranging from cloud-based to edge-computing-like deployments. In particular,} we propose to adopt the International Telecommunications Union (ITU) unified architecture for 5G and beyond. Based on the ITU's architecture, we provide insights on the main requirements and the major challenges of introducing ML to the multiple modalities of WLANs. \textcolor{blue}{Finally, we showcase the superiority of the architecture through an ML-enabled use case for future networks.}  %\textcolor{blue}{Finally, numerical results show the superiority of ML-based methods enabled by the architecture in front of currently adopted approaches.}
\end{abstract}

@@

ML is meant to empower a computational system for learning automatically, based on experience, so that future situations can be properly managed without having been programmed explicitly. Concerning wireless communications, there is a huge amount of unexploited data generated at both infrastructure and user levels, which could be extremely useful for learning complex patterns\textcolor{orange}{, which may allow improving network performance}. For instance, the behavior of users in a given network-oriented service can be predicted through ML, based on the information from previous sessions. \textcolor{orange}{Based on these predictions, network resources can be properly accommodated} in future sessions, thus leading to an optimized solution.

Unfortunately, the potential benefits of ML for real networks are currently limited by the existing infrastructure, which is not yet prepared to accommodate ML-oriented tasks such as data collection, processing, and output distribution. Instead, current networking systems are typically meant for delivering content, without taking into account the underlying characteristics of the processes that generate it.

@@

Apart from the ITU-T initiatives, other important standardization bodies such as the 3rd Generation Partnership Project (3GPP) or the European Telecommunications Standards Institute (ETSI) are currently working on the integration of data analytics to network functions. The 3GPP contemplates AI as one of the priority topics for shaping its upcoming release (Release 17), and architectural requirements are currently under study \cite{3gpp2019study}. Furthermore, we highlight the ETSI groups on Experiential Networked Intelligence (ENI) and Zero-touch network and Service Management (ZSM), which actively study the integration of AI to networks \cite{etsi2019architecture}. Unlike the ITU's unified architecture, most of the work held by the 3GPP and the ETSI focuses on centralized data collection and data analytics solutions. Nevertheless, we understand that the works in \cite{itu2019architecture, 3gpp2019study, etsi2019architecture} are complementary to each other.
% Alternative architecture: \cite{nekovee2019mobile}

@@

% Overview ML with formal definitions
To meet the abovementioned strict requirements, not only a technological innovation is expected (e.g., use of more spectrum or improve multiple-antennas technologies), but a paradigm shift is necessary when designing solutions for network planning, operation, and management. In particular, intelligent wireless networks need to be empowered with cognitive and context-aware capabilities, which may require additional infrastructure such as environmental sensors and cameras. To that end, ML is expected to be important during the lifetime of 5G and will become pervasive - as included from the beginning in their conception - in 6G networks.

@@

% 1 - Architectural proposals	
Apart from the specific ML solutions to problems in communications, some efforts have been made towards enabling AI-aware networking in more general terms. In this regard, several architectural proposals have been provided so far \cite{bi2015wireless,chih2017big,wang2018machine}. Most of the referenced works agree in the necessary steps for enabling big data analytics in cellular deployments: (1) data collection, (2) data preparation, (3) data analysis, and (4) decision making. Nevertheless, none of these works provide architectural guidelines to introduce ML to wireless networks. In contrast, the ITU's architecture looks deeper into the ML operation and targets the actual procedures involving information gathering, processing, and communication. In addition to the architectural guidelines, the ITU-T has provided a data handling framework for ML-aware networks \cite{itu2019data}, which defines processes concerning data collection, processing, and output distribution.

@@

\subsubsection{Cloud-Based User Association and Handover}
Most of the current user association and handover procedures held in WLANs typically rely on the strongest signal first (SSF) mechanism. This might be problematic in terms of load balancing \textcolor{orange}{and can potentially lead} to severe performance degradation \textcolor{orange}{in} saturated \textcolor{blue}{basic service sets (BSSs)}. By introducing ML, it is possible to handle contextual information such as the traffic load, which can \textcolor{orange}{be useful for} decision-making. Furthermore, mobility pattern prediction and predicted user requirements can be included in the system, thus empowering the association and handover mechanisms with insightful information. %\textcolor{orange}{Predictions can be provided, for instance, by Deep Neural Networks (DNN).}

@@

\subsubsection{Reinforcement Learning-Based Spatial Reuse} 
Spatial reuse aims to improve channel utilization through sensitivity adjustment mechanisms. However, selecting the best sensitivity threshold is not trivial, given the complex spatial interactions that occur among WLANs and the high action space. \textcolor{orange}{To that} purpose, reinforcement learning can be introduced to improve spectral efficiency in a decentralized manner so that each WLAN learns based on experience.

@@

The FG-ML5G was created in November 2017 by its parent group, the ITU-T Study Group 13, \textcolor{orange}{to study} the integration of ML mechanisms into future networks. This includes the definition of interfaces, protocols, data formats, and architectures. During its lifetime, the FG-ML5G has released several reports and contributions. Among them, we highlight the ITU's logical interoperable architecture for future networks \cite{itu2019architecture}, which defines an ML overlay that operates on the top of any unspecified underlay network technology (e.g., 3GPP, EdgeX, IEEE 802.11). The ITU's architecture aims to fulfill a set of technology-agnostic high-level requirements by the utilization of ML. For instance, the architecture must be able to support multiple types of data, thus taking advantage of heterogeneous data sources.

Figure \ref{fig:itu_ml_architecture} shows the elements that compose the ML overlay (management subsystem, multi-level ML pipeline, and closed-loop subsystem), which are further described in the following subsections. Based on this \textcolor{orange}{standard} overlay, ML applications can be instantiated in the logical entities (represented by white boxes).

\subsection{Management Subsystem} 
The management subsystem is in charge of the deployment and the orchestration of the ML services \textcolor{orange}{that operate} in the underlying network. To that purpose, the \textcolor{blue}{machine learning function orchestrator (MLFO)} entity is defined. \textcolor{blue}{The MLFO is first instantiated by a declarative intent that uses a meta language, which specifies the ML use case to be applied, including initialization, policies, and constraints. Then, the MLFO initializes the elements of the multi-level ML pipeline in the corresponding devices and monitors their operation during execution.}

@@

@@

Figure \ref{fig:overview_learning_approaches} illustrates the enterprise and residential-like deployments, as well as a set of IEEE 802.11 mechanisms that can \textcolor{orange}{facilitate the adoption of the ML-based architecture in WLANs.} \textcolor{blue}{These mechanisms, which can be used for data collection, training, and output distribution, are as follows:}
\textcolor{blue}{\begin{itemize}
	\item Information gathering (802.11k/r/v): a given ML mechanism can use information about the network topology or RF measurements to infer the behavior of other devices or to extract important environmental characteristics.
	\item Interoperability (802.11f/u): interoperability with other networks can be used to perform coordinated operations (e.g., scheduling, resource allocation). Besides, inter-AP communication procedures can enable centralized/coordinated mechanisms (e.g., federated learning).
	\item Security (802.11w): ML mechanisms can use management frames that are protected so that a higher level of security is granted.
	\item Validation (802.11t): Performance evaluation in WLANs through test metrics can be of great utility to define optimization goals within the ML  operation.
\end{itemize}}

%\begin{table*}[t!]
%	\caption{Features supported by 802.11 amendments that can enable the introduction of ML methods to WLANs.}
%	\label{tab:opportunities_amendments}
%	\centering
%	\begin{tabular}{|p{.14\textwidth}|p{.1\textwidth}|p{.6\textwidth}|}
%		\hline
%		\textbf{Feature} & \textbf{Amendments} & \textbf{Opportunities for ML application} \\\hline
%		Information gathering & 802.11k/r/v & A given ML mechanism can use information about the network topology or RF measurements to infer the behavior of other devices or to extract important environmental characteristics.\\\hline
%		Interoperability & 802.11f/u & Interoperability with other networks can be used to perform coordinated operations (e.g., scheduling, resource allocation). Besides, inter-AP communication procedures can enable centralized/coordinated mechanisms (e.g., federated learning). \\\hline
%		Security & 802.11w & ML mechanisms can use management frames that are protected so that a higher level of security is granted.\\\hline
%		Validation & 802.11t & Performance evaluation in WLANs through test metrics can be of great utility to define optimization goals within the ML  operation.\\\hline
%	\end{tabular}
%\end{table*}

%%% CHALLENGES IN IEEE 802.11 WLANs
\subsection{Challenges in IEEE 802.11 WLANs}
\label{section:ieee_80211_wlans}
The application of ML methods in WLANs is tightly tied to the technological challenges posed by these types of networks. The major challenges encountered in wireless communications stand for fast data expiry and lack of resources for data handling (e.g., storage, computation, information exchange). Regarding Wi-Fi networks, we find the following challenges:
\begin{enumerate}
	\item \textbf{Network dynamics:} channel fluctuations (e.g., due to multipath fading), users mobility and varying traffic needs entail a big challenge to ML applications. As a result of network dynamics, ML methods need to constantly adapt to the environment, which may be achieved through continuous re-training.
	\item \textbf{Limited communication resources:} since Wi-Fi operates under unlicensed bands, resources are scarce and shared. Thus, any potential communication required by a certain ML mechanism (as for distributed learning) may fail or be delayed if the medium is congested. As a result, the ML operation must be robust and resilient enough to react to potential communication issues.
	\item \textbf{Limited computation and storage resources:} computation and storage resources may also be limited in WLANs, especially in residential-like deployments. Therefore, the ML operation should include computation-efficient procedures. Another implication of limited resources lies in the availability of information to be used by ML algorithms, especially for online learning methods.
	\item \textbf{Adversarial environment:} in many cases, Wi-Fi deployments are chaotic in the sense that many overlapping \textcolor{orange}{BSSs} coexist without any kind of cooperation. This is a particularly interesting challenge for ML \textcolor{orange}{methods}, which may lead to adversarial settings where different agents compete for the same resources. Moreover, multi-vendor devices may implement different ML mechanisms that may lead to clashing interests.
	\item \textbf{Legacy devices:} \textcolor{orange}{similarly} to the previous case, ML-empowered WLANs may coexist with other legacy networks that do not perform any intelligent operation. It is then required for ML methods to be aware of those legacy devices, so that \textcolor{orange}{unfair} situations are avoided.
\end{enumerate}

@@

In edge-oriented mechanisms, the ML operation is mainly ruled by edge devices (e.g., APs and/or STAs), which, contrary to the cloud approach, typically lack powerful computation and storage resources. In consequence, edge-oriented mechanisms may only allow using simple and light-weight computing ML algorithms. Nevertheless, edge servers can be placed for providing more powerful solutions \textcolor{orange}{promptly}. The edge-oriented approach is useful for real-time ML applications that manage local (and even highly-varying) information.

Besides cloud and edge-oriented settings, we may distinguish between methods based on \textcolor{orange}{their cooperation degree}. In cooperative approaches, nodes interact among them for the sake of conducting the learning operation \textcolor{orange}{jointly} (e.g., use a shared reward). With this aim, reliable and timely connections among learners are often required. In this regard, \cite{lin2017deep} showed the role of communications on speeding up a distributed training procedure over a set of nodes in a network. Alternatively, for the non-cooperative case, the learning operation may lead to adversarial settings, especially since WLANs share resources such as frequency channels.

@@

Specifically, the training procedure consists of the following steps (shown \textcolor{orange}{in} red):
\begin{enumerate}
	\item \textbf{Data collection:} the edge server collects information of different \textcolor{orange}{kind} from either APs and STAs, such as user information (e.g., current location), performance (e.g., throughput, delay), application data (e.g., traffic load), or channel status reports (e.g., the power received from interfering nodes). This information can be used either for training or feeding auxiliary algorithms that help the main AP association procedure (e.g., predict user behavior).
	\item \textbf{ML input preparation:} the data collected at the cloud is prepared by the pre-processor so that the ML method can properly manage it. For instance, in case of applying a multiple linear regression, the input user information needs to be converted into normalized features (e.g., convert the throughput given in Mbps into a scalar between 0 and 1).
	\item \textbf{ML method application:} \textcolor{orange}{the final output of the ML takes policies into account. Accordingly,} a given AP may \textcolor{orange}{consider} specific constraints regarding its occupation or the maximum number of associated STAs. The policies are strongly tied to the capabilities of the devices or the existing regulations (e.g., frequency channels available, maximum regulated transmission power).
	\item \textbf{Output distribution:} once the ML method generates the output (i.e., the predicted function for new (re)associations), it is distributed throughout the sink edge servers, which are then prepared to give quick response to new cases.
\end{enumerate}

@@

\textcolor{blue}{To showcase the potential of the ML-based architecture through numerical results,\footnote{\textcolor{blue}{Given the novelty of the technologies studied in this paper, our results have been obtained from well-know models, hence their accuracy is tied to them. Nevertheless, this is a first step to understand the potential benefits of using an ML-based architecture in next-generation wireless networks. For the sake of reproducibility and disclosure, all the source code is open and publicly available at \url{https://github.com/fwilhelmi/machine_learning_aware_architecture_wlans}, accessed on 20 January 2020.}} Fig. \ref{fig:results_use_case} compares the performance of classical AP association procedure (strongest signal first (SSF)) with a novel ML-based approach (based on vanilla neural networks). As illustrated, the ML approach allows capturing complex patterns from dense deployments, which is useful to improve satisfaction ratio and fairness in the AP (re)association problem.}

\begin{figure}[ht!]
	\centering
	\includegraphics[width=0.9\columnwidth]{results_use_case}
	\caption{\textcolor{blue}{Results of applying the hybrid ML-based AP (re)association and handover in IEEE 802.11 WLANs. Different number of STAs are considered to showcase the impact of load on APs. For each mechanism, the average performance is represented by a star.}}
	\label{fig:results_use_case}
\end{figure}

@@

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://mirror.ctan.org/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
\bibliographystyle{IEEEtran}
\bibliography{bibliography}

@@

@@

\vfill
